Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-05-24 ms.qmd
Updates

## Abstract
Behavioural flexibility is crucial for animals adapting to and colonizing new environments. However, novel environmental conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how invasive populations establish in new locations. Given the advantages associated with behaving flexibly during the invasion process we predict that behavioural flexibility in invasive species should be less susceptible to the impacts of early environmental conditions. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_, with distinct histories of invasion. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. Contrary to our expectations, we found no significant differences in the reversal learning rate between treatments or species. Behavioural flexibility appears to be robust to prenatal temperature and CORT which may be one reason why _L. delicata_ has been so successful as an invasive species. However, the lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that other processes are likely more influential in invasion success than behavioural flexibility itself.

@@

The conditions in which the organisms develop can be altered by the circumstances faced by the parents. For instance, exposing threespine sticklebacks females to a model predator while yolking eggs affected offspring tedency to use social cues [@feng2015effect]. Under stressful situations, animal's sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. For example, human disturbance alters snapping turtles' nest-site choice, ultimately impacting the temperature at which eggs are incubated [@kolbe2002impact]. This way, parental effects like transmitted GCs, are predicted to interact with other factors from the early environment to influence the offspring phenotype. Both GCs and temperature are known to exert potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic], but little is known about how their ineraction could affect behavioural flexibility. Since behaving flexibly is essential for facing novel circumstances, studying the interactive effect of early environment on behavioural flexibility is crucial if we aim to understand animals' capability to adapt to environmental changes.

@@

This study aimed to determine the interactive effects of prenatal temperature and GCs on behavioural flexibility in two closely related species, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_).

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. In contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], we did not find any effect of CORT on the speed at which lizards learn. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. In some systems, exposure to CORT during development has hormetic effects on learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. The CORT treatment employed here is known to impact other aspects such as growth or mortality in these species (unpublished data), but the effect of the same dose can differ between tissues (@lemaire_prenatal_2000). More striking is the absence of an effect of incubation temperature since most of the studies in reptiles reported better performance when incubated at higher temperatures in different associative learning and motor tasks [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. We know that our prenatal treatments affect other traits in both species of lizards used in this experiment (unpublished data). However, it is possible that the treatments affected tissues differently. An alternative hypothesis is that the effects of the prenatal treatments are age-dependent [see @frisone2002social]. Here, the lizards' age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_, and when this variable was included in the models, we did no find any significant effect of age (see Supplementary material). However, the treatments may have consequences at earlier or later stages of life that are not being captured in our experiment. Conversely, the effects of the early environment may likely vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. In addition, prenatal GCs can have programmatic effects on HPA axis sensitivity, potentially impacting the lizards' sensitivity to stressors [@crino_corticosterone_2014]. This altered stress response may influence learning dynamics through hormetic effects of CORT [@du_dynamic_2009], particularly in stressful contexts [@taborsky2021towards]; although such effects were not likely observable in our experimental design. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks and at different ages, testing the effect of stress during the learning process, and measuring the neurophysiological effects on different brain nuclei.

Despite sharing life-history traits, only _L. delicata_ has proven successful in colonising various global regions [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], and we predicted this may relate to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. Nevertheless, we did not find significant differences between treatments or species when we compared their performance in a reversal task. We found that, on average, all the lizards from the four treatments and the two species learned the reversal task - indicated by estimated slopes that are significantly different from zero - and that they learn, on average, at the same rate. Furthermore, these results were consistent when only those individuals who performed better in the previous associative task were included (see Supplementary material).

Previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, Bezzina _et al._ [@bezzina2014does] found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species. In addition, we did not find any effect of the treatments on either species. Both results contrast our original predictions and may indicate that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than behavioural flexibility. For instance, the mentioned differences in exploratory behaviour can be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas, since exploring more can also enhance the ability to effectively locate novel resources in _L. delicata_ [@chapple_know_2011].

In conclusion, our results indicate that lizards performed the reversal learning task, a measure of behavioural flexibility, at similar rates, with no discernible differences between treatments or species. These findings contrast our initial predictions and suggest that other aspects of _L. delicata_ biology may be more influential for their success as invaders than behavioural flexibility alone. The present study also indicates that prenatal CORT and temperature do not affect learning abilities in these species. Nevertheless, it remains possible that the effects of the early environment are task-dependent, and future studies should explore performance across different cognitive tasks and the effects of both treatments on brain function.

@@

Results for _L. delicata_:
```{r, tbl-agedeli, results='asis', out.widht = "26cm"}
#| label: tbl-agedeli
#| tbl-cap: "Summary model_age_deli"
deli_age_res <- summary(model_age_deli)
# Extract fixed effects names and their corresponding estimates, CIs, etc.
fixed_effects <- deli_age_res$fixed %>%
  mutate_all(~ format_dec(., 2))
Predictors <- rownames(fixed_effects)
deli_age_tbl_df <- cbind(Predictors, fixed_effects)
# Convert the fixed effects data frame to a flextable
flextable(deli_age_tbl_df) %>%
  set_table_properties(., width = 1) %>%
  fontsize(size = 10, part = "all")
```

@@

Results for _L. guichenoti_:
```{r, tbl-ageguich, results='asis'}
#| label: tbl-ageguich
#| tbl-cap: "Summary model_age_guich"
guich_age_res <- summary(model_age_guich)
# Extract fixed effects names and their corresponding estimates, CIs, etc.
fixed_effects <- guich_age_res$fixed %>%
  mutate_all(~ format_dec(., 2))
Predictors <- rownames(fixed_effects)
guich_age_tbl_df <- cbind(Predictors, fixed_effects)
# Convert the fixed effects data frame to a flextable
flextable(guich_age_tbl_df)%>%
  set_table_properties(., width = 1) %>%
  fontsize(size = 10, part = "all")
```

>>>>

## Abstract
Behavioural flexibility is crucial for animals adapting to environmental changes. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on animals' cognition through the transmission of glucocorticoids - stress-related hormones. At the same time, stressful conditions can influence parental behaviour during nesting and, in consequence the developmental conditions such as temperature. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures to impair with lizards' performance in the reversal task. Also, given _L. delicata_ success as an invassive species, we expected this species to behave more flexible and be less affected by the incubation conditions. Contrary to our expectations, we found no significant differences in the reversal learning rate between treatments or species. Behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. Furthermore, the lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that other processes are likely more influential in invasion success than behavioural flexibility itself.

@@

The conditions in which the organisms develop can be altered by the circumstances faced by the parents. For instance, exposing threespine sticklebacks females to a model predator while yolking eggs, affected offspring tedency to use social cues [@feng2015effect]. Under stressful situations, animal's sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. For example, human disturbance alters snapping turtles' nest-site choice, ultimately impacting the temperature at which eggs are incubated [@kolbe2002impact]. Because animals' responses to abrupt changes in temperature are mediated by GCs [@Crino_2023], and this response can affect animals' decision-making while nesting, GCs transmission and early therma environment are predicted to interact to shape offspring phenotype. In this sense, both GCs and temperature are known to exert potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic], but little is known about how their ineraction could affect behavioural flexibility. Since behaving flexibly is essential for facing novel circumstances, studying the interactive effect of early environment on behavioural flexibility is crucial if we aim to understand animals' capability to adapt to environmental changes.

@@

This study aimed to determine the interactive effects of prenatal temperature and GCs on behavioural flexibility in two closely related species, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We expected incubation conditions to influence behavioural flexibility in both species; and we also predicted that _L. delicata_ would be more flexible than _L. guichenoti_ and less affected by the early environment. Nevertheless, we did not find significant differences between treatments or species when we compared their performance in a reversal task. Our results show that, on average, all the lizards from the four treatments and the two species learned the reversal task - indicated by estimated slopes that are significantly different from zero - and that they learn, on average, at the same rate. Furthermore, these results were consistent when only those individuals who performed better in the previous associative task were included (see Supplementary material).

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. In contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], we did not find any effect of CORT on the speed at which lizards learn. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016].  More striking is the absence of an effect of incubation temperature since most of the studies reported better performance in different associative learning and motor tasks when reptiles were incubated at higher temperatures [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards to learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. Our results suggest that behavioural flexibility, measured here as lizards' ability to solve a reversal learning task, is robust to prenatal CORT and temperature in both species of skinks. One potential limitation of our study is whether the dose we applied onto eggs increased CORT levels enough to impact lizards, as exposure to CORT during development has hormetic effects on learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. We find this possibility unlikely since in a previous experiment, we observed that a similar dose affected other traits such as growth and mortality in _L. delicata_ (unpublished data), but that is a scenario that cannot be dismissed. Furthermore, the effect of the same dose can differ between tissues (@lemaire_prenatal_2000), affecting mortality and growth, but nor mortality. Another alternative hypothesis is that the effects of the prenatal treatments are age-dependent [see @frisone2002social]. Here, the lizards' age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_, and when this variable was included in the models, we did no find any significant effect of age (see Supplementary material). However, the treatments may have consequences at earlier or later stages of life that are not being captured in our experiment. Conversely, the effects of the early environment may likely vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. In addition, prenatal GCs can have programmatic effects on HPA axis sensitivity, potentially impacting the lizards' sensitivity to stressors [@crino_corticosterone_2014]. This altered stress response may influence learning dynamics through hormetic effects of CORT [@du_dynamic_2009], particularly in stressful contexts [@taborsky2021towards]; although such effects were not likely observable in our experimental design. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks and at different ages, testing the effect of stress during the learning process, and measuring the neurophysiological effects on different brain nuclei.

Despite sharing life-history traits, only _L. delicata_ has proven successful in colonising various global regions [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], and we predicted this may relate to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. Nevertheless, we did not find differences in reversal learning between _L. delicata_ and _L. guichenoti_, and in both species, behavioural flexibility seems robust to the influence of prenatal conditions. Previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, Bezzina _et al._ [@bezzina2014does] found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species. In addition, we did not find any effect of the treatments on either species. Both results contrast our original predictions and may indicate that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than behavioural flexibility. For instance, the mentioned differences in exploratory behaviour can be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas, since exploring more can also enhance the ability to effectively locate novel resources in _L. delicata_ [@chapple_know_2011].

In conclusion, our results indicate that lizards performed the reversal learning task, a measure of behavioural flexibility, at similar rates, with no discernible differences between treatments or species. These findings contrast our initial predictions and suggest that behavioural flexibility in both species is robust against conditions during early stages of development. Furthermore, our results indicate that other aspects of the biology of _L. delicata_ may be more influential for their success as invaders than behavioural flexibility alone. Nevertheless, it remains possible that the effects of the prenatal CORT and temperature on cognition depend on other aspects like the age at which animals are tested or the cognitive task employed; and future studies should explore performance across different cognitive tasks and the effects of both treatments on brain function.

@@

Results for _L. delicata_ when the model included the age:
```{r, tbl-agedeli, results='asis', out.widht = "26cm"}
#| label: tbl-agedeli
#| tbl-cap: "Summary model_age_deli"
deli_age_res <- summary(model_age_deli)
# Extract fixed effects names and their corresponding estimates, CIs, etc.
fixed_effects <- deli_age_res$fixed %>%
  mutate_all(~ format_dec(., 2))
Predictors <- rownames(fixed_effects)
deli_age_tbl_df <- cbind(Predictors, fixed_effects)
# Convert the fixed effects data frame to a flextable
flextable(deli_age_tbl_df) %>%
  set_table_properties(., width = 1) %>%
  fontsize(size = 10, part = "all")
```

@@

Results for _L. guichenoti_ when the model included the age:
```{r, tbl-ageguich, results='asis'}
#| label: tbl-ageguich
#| tbl-cap: "Summary model_age_guich"
guich_age_res <- summary(model_age_guich)
# Extract fixed effects names and their corresponding estimates, CIs, etc.
fixed_effects <- guich_age_res$fixed %>%
  mutate_all(~ format_dec(., 2))
Predictors <- rownames(fixed_effects)
guich_age_tbl_df <- cbind(Predictors, fixed_effects)
# Convert the fixed effects data frame to a flextable
flextable(guich_age_tbl_df)%>%
  set_table_properties(., width = 1) %>%
  fontsize(size = 10, part = "all")
```
